Rising from the Ashes II: The Bar-driven Abundance Bimodality of the Milky Way

### 2025-02-19 paper/ms.tex
Change from global to Z-dependent SFR

\newcommand{\Msun}{\ensuremath{M_{\odot}}}
\newcommand{\Gyr}{\ensuremath{\textrm{Gyr}}}
\newcommand{\Myr}{\ensuremath{\textrm{Myr}}}
\newcommand{\yr}{\ensuremath{\textrm{yr}}}
\newcommand{\kpc}{\ensuremath{\textrm{kpc}}}
\newcommand{\ckpc}{\ensuremath{\textrm{ckpc}}}
\newcommand{\pc}{\ensuremath{\textrm{pc}}}
\newcommand{\kms}{\ensuremath{\textrm{km}/\textrm{s}}}
\newcommand{\tocite}{\textcolor{blue}{cite}}
\newcommand{\FeH}{\ensuremath{[\textrm{Fe}/\textrm{H}]}}
\newcommand{\MgFe}{\ensuremath{[\textrm{Mg}/\textrm{Fe}]}}
\newcommand{\OFe}{\ensuremath{[\textrm{O}/\textrm{Fe}]}}
\newcommand{\alphaFe}{\ensuremath{[\alpha/\textrm{Fe}]}}
\newcommand{\tform}{\ensuremath{t_{\textrm{form}}}}
\newcommand{\dex}{\ensuremath{\textrm{dex}}}
\newcommand{\Msunyr}{\ensuremath{\Msun/\textrm{yr}}}
\newcommand{\Atmax}{\ensuremath{A_{2,\textrm{max}}}}

@@

In \citet{2024arXiv240707985B} (hereafter Paper~I), we proposed an alternate scenario for the formation of the bimodality, driven by a brief ($\sim300\,\Myr$) quiescent period in the Galaxy's history (an observational claim with a longer period of quiescence was made in \citet{2016A&A...589A..66H}). This period has two effects. First, the reduced SFR lowers the number of $\alpha$-producing Type~II SNe relative to Type~Ia\footnote{These SNe occur on different timescales (10s of Myr after star formation for Type~II, as compared to 100s of Myr to Gyrs for Type~Ia), so if the SFR drops, the number of Type~II SNe drops faster than Type~Ia.}, causing a sharp decline in the \alphaFe{} ratio in the star-forming gas. Second, the lower SFR results in fewer stars forming in the intermediate region between the high- and low-$\alpha$ sequences, reducing the occurrence of this transitional population in present-day observations. The combination of these two effects creates the valley between the high- and low-$\alpha$ sequences. This mechanism resembles two-phase infall models that incorporate a temporary halt in star formation \citep[][and references therein]{2024arXiv240511025S}, though the quiescent period in Paper~I is much shorter.

@@

We select subhalo 172175 (the Subfind ID at snapshot 40) for its resemblance to the Milky Way. We then studied the main descendant of this subhalo at $z=0$ (subhalo 392276 at snapshot 99). A summary of its key properties is given in Table~\ref{tab:summ}.

@@

\subsection{Observations}\label{ssec:obs}
We make use of two observational data sets. First, we use the stellar abundances from the ASPCAP pipeline of APOGEE DR17 \citep[][J.A.~Holtzman et al., in preparation]{2016AJ....151..144G,2017AJ....154...28B,2017AJ....154...94M,2022ApJS..259...35A}. We make the same selection cuts as in Section~2.4 of Paper I. These criteria select giants with high quality abundance measurements and angular momenta similar to the Sun's. This results in a sample of 54,777 stars. We use Fe to track total metallicity and Mg alone as an $\alpha$-element.

@@

To understand the key events driving the behavior around the dashed line in Figure~\ref{fig:alpha}, we examine the evolution of several properties of our galaxy in Figure~\ref{fig:history}: its SFH, the BH accretion rate, and the growth of the virial mass. The vertical dashed line in each panel marks the transition at $t\sim3.2\,\Gyr$ from the high- to low-$\alpha$ sequences, as in Figure~\ref{fig:alpha}. The upper panel in the left column shows the the SFR (computed for all gas cells in the galaxy). There are two peaks at $t\sim2.5\,\Gyr$ and $t\sim4.5\,\Gyr$ with maximum values of $50\,\Msun/\textrm{yr}$ and $30\,\Msun/\textrm{yr}$, respectively. Around the high- to low-$\alpha$ transition, there is a dip in the SFR, which drops by an order of magnitude to about $3\,\Msun/\textrm{yr}$. The right panel zooms in on the period between $t=2\,\Gyr$ and $5\,\Gyr$, where we observe a sharp recovery in the SFR following the quiescent phase. In the span of a single snapshot (roughly $150\,\Myr$), the SFR increases from about $3\,\Msun/\textrm{yr}$ to $10\,\Msun/\textrm{yr}$.

@@

@@

This figure shows the following sequence of events:
\begin{enumerate}
    \item Bar forms: A steady increase in the bar strength, as indicated by $A_2/A_0$ for star particles with $R<2\,\kpc$, from $\sim0.05$ to $0.4$ starting around $2.8\,\textrm{Gyr}$. This rise is accompanied by the appearance of elongated features in the gas and stars consistent with a bar.
    \item BH accretion increases: Following the increase in bar strength by about a snapshot ($\sim150\,\Myr$ here), the BH accretion rate ($\dot{M}_{\textrm{BH}}/\dot{M}_{\textrm{edd}}$) shows a significant spike, rising from a minimum of $\sim0.08$ at $t=2.84\,\Gyr$ to a maximum of $\sim0.29$ at $t=3.13\,\Gyr$ for one snapshot.
    \item SFR declines: The SFR declines starting from a maximum of $54.3\,\Msunyr$ at $t=2.84\,\Gyr$ down to $3.6\,\Msunyr$ at $t=3.28\,\Gyr$. In the next snapshot at $t=3.45\,\Gyr$ the SFR recovers to $12.2\,\Msunyr$. Figure~\ref{fig:history} shows that it reaches its second maximum of $30.9\,\Msunyr$ at $4.5\,\Gyr$.
\end{enumerate}

@@

\subsection{Cause of Bimodality}\label{ssec:bim_cause}
We argue that the available evidence for the cause of the bimodality aligns with the scenario outlined in Paper~I. That study presented an idealized simulation resembling the merger between the Milky Way and GSE, where the orbital parameters were varied slightly across a grid of 27 simulations. The results showed that simulations featuring a brief quiescent period produced a bimodal abundance distribution. Two conditions for the bimodality to arise have to be met: (1) a declining \alphaFe{} with time (here done through our post-processing step, see Figure~\ref{fig:alpha}), and (2) a gap in SFR. In Paper~I, the gap in SFR was due to a merger, but here we argue it is due to the formation of a bar.

@@

During this period of suppressed SF, we argue that the lack of enrichment of Type~II relative to Type~Ia SNe leads to a lower rate of Mg production. The typical lifetime of a Type~II SN progenitor in this model is $\sim40\,\Myr$ \citep{2018MNRAS.473.4077P}, and so the hundreds of Myr of suppressed SF is short enough to restrict the production of $\alpha$-elements. Therefore, \MgFe{} rapidly declines. This, combined with a lack of SF in the first place, leads to a scarcity of star particles in the intermediate region between the high- and low-$\alpha$ sequences. A more in-depth explanation is given in Section~4.1 in Paper~I.

@@

Notably, both the fiducial and $\alpha$-enhanced galaxy show a slight rebound effect in \MgFe{}. The star particles which form directly after the dashed line when the SFR has just recovered have a slightly lower \MgFe{} than stars which form later on, by about $0.1$ to $0.2\,\dex$. This was also predicted in Figure~9 of Paper~I, where it was argued that during the period of suppressed star formation, the \alphaFe{} ratio of the star-forming gas drops sharply due to the reduced contribution of Type~II relative to Type~Ia SNe. Later, the \alphaFe{} of the gas will recover when the SFR also recovers, but there is a brief window when old, low-$\alpha$ stars can form. A similar behavior was seen in the one-zone models with bursty SFHs in \citet{2020MNRAS.498.1364J}.

@@

\subsection{Cause of Quiescence}\label{ssec:cause_qui}
In Paper~I, AGN activity from a merger was the suspected cause for the quiescent period. In our galaxy here, there is indeed a brief burst in AGN accretion at the time of the merger (middle panel of Figure~\ref{fig:history}). Based on this burst, it is also reasonable to suspect that AGN activity is also responsible for the quiescent period in our galaxy. However, we argued in Section~\ref{ssec:sequence_of_events} that the cause of the localized spike in the BH accretion rate is not due to a merger but instead due to the formation/strengthening of a bar. \edit1{This connection is further supported by estimates that the Milky Way's bar formed $\sim8\,\Gyr$ ago \citep[e.g.,][]{2019MNRAS.490.4740B,2024MNRAS.530.2972S}, roughly concurrent with the formation of the bimodality \citep{2013A&A...560A.109H,2023MNRAS.525.2208R,2024MNRAS.535..392L}.}

@@

The formation of the bimodality, when the galaxy transition from producing high- to low-\MgFe{} star particles (Figure~\ref{fig:alpha}), is coincident with a global supression of star formation (Figure~\ref{fig:history}). This suppression of star formation is preceded by the formation of a bar and subsequent AGN activity (Figure~\ref{fig:seq}). \edit1{This scenario is plausible for the Milky Way, as recent estimates indicate that the bar formed around $8\,\Gyr$ ago \citep{2019MNRAS.490.4740B,2024MNRAS.530.2972S}, coinciding with the onset of the bimodality.}

We argue that the reduction in the SFR has two effects which lead to the $\alpha$-bimodality. First, a lower production of Mg relative to Fe is the natural result of a lack of Type~II relative to Type~Ia SNe. The latter remain high due to their longer timescales, expediting a drop in \MgFe{}. Second, the lack of SF means that stars at \MgFe{} intermediate between the high- and low-$\alpha$ sequences never form. This argument is the same as presented in Paper~I and is consistent with the one-zone chemical evolution models of \citet{2020MNRAS.498.1364J}. \edit1{Because of observational errors in age, it is not possible to make a direct comparison between the simulated galaxy and datae (Figure~\ref{fig:alpha} and Section~\ref{ssec:compare_obs}).}

This work adds further support to a scenario in which a quiescent period in the Milky Way's past is a plausible explanation for the Milky Way's abundance bimodality. We argued in Paper~I that the GSE merger could trigger this period. In this work we have argued that the formation of the Milky Way's bar could be responsible. Regardless of this scenario's relevance to the Milky Way, we also predict that the presence of a bar in external galaxies is correlated with the presence of an $\alpha$-bimodality.

@@

\figsetstart
\figsetnum{7}
\figsettitle{Abundance plane for a random subset of galaxies in our Milky Way-progenitor mass sample, with and without $\alpha$-enhancement.}
\figsetgrpstart
\figsetgrpnum{7.1}
\figsetgrptitle{Subhalo 392276.}
\figsetplot{app_392276.pdf}
\figsetgrpnote{Abundance plane for subhalo 392276.}
\figsetgrpend
\figsetgrpstart
\figsetgrpnum{7.2}
\figsetgrptitle{Subhalo 2.}
\figsetplot{app_2.pdf}
\figsetgrpnote{Abundance plane for subhalo 2.}
\figsetgrpend
\figsetgrpstart
\figsetgrpnum{7.3}
\figsetgrptitle{Subhalo 10.}
\figsetplot{app_10.pdf}
\figsetgrpnote{Abundance plane for subhalo 10.}
\figsetgrpend
\figsetgrpstart
\figsetgrpnum{7.4}
\figsetgrptitle{Subhalo 143882.}
\figsetplot{app_143882.pdf}
\figsetgrpnote{Abundance plane for subhalo 143882.}
\figsetgrpend
\figsetgrpstart
\figsetgrpnum{7.5}
\figsetgrptitle{Subhalo 167392.}
\figsetplot{app_167392.pdf}
\figsetgrpnote{Abundance plane for subhalo 167392.}
\figsetgrpend
\figsetgrpstart
\figsetgrpnum{7.6}
\figsetgrptitle{Subhalo 289388.}
\figsetplot{app_289388.pdf}
\figsetgrpnote{Abundance plane for subhalo 289388.}
\figsetgrpend
\figsetgrpstart
\figsetgrpnum{7.7}
\figsetgrptitle{Subhalo 300903.}
\figsetplot{app_300903.pdf}
\figsetgrpnote{Abundance plane for subhalo 300903.}
\figsetgrpend
\figsetgrpstart
\figsetgrpnum{7.8}
\figsetgrptitle{Subhalo 348901.}
\figsetplot{app_348901.pdf}
\figsetgrpnote{Abundance plane for subhalo 348901.}
\figsetgrpend
\figsetgrpstart
\figsetgrpnum{7.9}
\figsetgrptitle{Subhalo 398784.}
\figsetplot{app_398784.pdf}
\figsetgrpnote{Abundance plane for subhalo 398784.}
\figsetgrpend
\figsetgrpstart
\figsetgrpnum{7.10}
\figsetgrptitle{Subhalo 404818.}
\figsetplot{app_404818.pdf}
\figsetgrpnote{Abundance plane for subhalo 404818.}
\figsetgrpend
\figsetgrpstart
\figsetgrpnum{7.11}
\figsetgrptitle{Subhalo 425719.}
\figsetplot{app_425719.pdf}
\figsetgrpnote{Abundance plane for subhalo 425719.}
\figsetgrpend
\figsetgrpstart
\figsetgrpnum{7.12}
\figsetgrptitle{Subhalo 439099.}
\figsetplot{app_439099.pdf}
\figsetgrpnote{Abundance plane for subhalo 439099.}
\figsetgrpend
\figsetgrpstart
\figsetgrpnum{7.13}
\figsetgrptitle{Subhalo 465255.}
\figsetplot{app_465255.pdf}
\figsetgrpnote{Abundance plane for subhalo 465255.}
\figsetgrpend
\figsetgrpstart
\figsetgrpnum{7.14}
\figsetgrptitle{Subhalo 494709.}
\figsetplot{app_494709.pdf}
\figsetgrpnote{Abundance plane for subhalo 494709.}
\figsetgrpend
\figsetgrpstart
\figsetgrpnum{7.15}
\figsetgrptitle{Subhalo 510273.}
\figsetplot{app_510273.pdf}
\figsetgrpnote{Abundance plane for subhalo 510273.}
\figsetgrpend
\figsetgrpstart
\figsetgrpnum{7.16}
\figsetgrptitle{Subhalo 547293.}
\figsetplot{app_547293.pdf}
\figsetgrpnote{Abundance plane for subhalo 547293.}
\figsetgrpend
\figsetgrpstart
\figsetgrpnum{7.17å}
\figsetgrptitle{Subhalo 576516.}
\figsetplot{app_576516.pdf}
\figsetgrpnote{Abundance plane for subhalo 576516.}
\figsetgrpend
\figsetend

>>>>

\newcommand{\Msun}{\ensuremath{M_{\odot}}}
\newcommand{\Gyr}{\ensuremath{\textrm{Gyr}}}
\newcommand{\Myr}{\ensuremath{\textrm{Myr}}}
\newcommand{\yr}{\ensuremath{\textrm{yr}}}
\newcommand{\kpc}{\ensuremath{\textrm{kpc}}}
\newcommand{\ckpc}{\ensuremath{\textrm{ckpc}}}
\newcommand{\pc}{\ensuremath{\textrm{pc}}}
\newcommand{\kms}{\ensuremath{\textrm{km}/\textrm{s}}}
\newcommand{\tocite}{\textcolor{blue}{cite}}
\newcommand{\FeH}{\ensuremath{[\textrm{Fe}/\textrm{H}]}}
\newcommand{\MgFe}{\ensuremath{[\textrm{Mg}/\textrm{Fe}]}}
\newcommand{\OFe}{\ensuremath{[\textrm{O}/\textrm{Fe}]}}
\newcommand{\alphaFe}{\ensuremath{[\alpha/\textrm{Fe}]}}
\newcommand{\tform}{\ensuremath{t_{\textrm{form}}}}
\newcommand{\dex}{\ensuremath{\textrm{dex}}}
\newcommand{\Msunyr}{\ensuremath{\Msun/\textrm{yr}}}
\newcommand{\Msunyrdex}{\ensuremath{\Msun/\textrm{yr}/\textrm{dex}}}
\newcommand{\Atmax}{\ensuremath{A_{2,\textrm{max}}}}

@@

In \citet{2024arXiv240707985B} (hereafter Paper~I), we proposed an alternate scenario for the formation of the bimodality, driven by a brief ($\sim300\,\Myr$) quiescent period in the Galaxy's history \edit1{in a narrow metallicity bin, assuming the gas-phase \alphaFe{} is declining sufficiently rapidly.}\footnote{An observational claim with a longer period of global quiescence was made in \citet{2016A&A...589A..66H}.} The lower SFR results in fewer stars forming in the intermediate region between the high- and low-$\alpha$ sequences, reducing the occurrence of this transitional population in present-day observations. \edit1{A global quiescent period is sufficient for producing such gaps in the metallicity-dependent SFR, though it is not necessary.} This mechanism resembles two-phase infall models that incorporate a temporary halt in star formation \citep[][and references therein]{2024arXiv240511025S}, though the quiescent period in Paper~I is much shorter.

@@

\edit1{We used \texttt{FOF} and \texttt{SUBFIND} in order to identify substructure \citep{2005Natur.435..629S,2009MNRAS.399..497D}.} We select subhalo 172175 (the \texttt{SUBFIND} ID at snapshot 40) for its resemblance to the Milky Way. We then studied the main descendant of this subhalo at $z=0$ (subhalo 392276 at snapshot 99). A summary of its key properties is given in Table~\ref{tab:summ}.

@@

\subsection{Observational Data}\label{ssec:obs}
We make use of two observational data sets. First, we use the stellar abundances from the ASPCAP pipeline of APOGEE DR17 \citep[][J.A.~Holtzman et al., in preparation]{2016AJ....151..144G,2017AJ....154...28B,2017AJ....154...94M,2022ApJS..259...35A}. We make the same selection cuts as in Section~2.4 of Paper I. These criteria select giants with high quality abundance measurements and angular momenta similar to the Sun's. This results in a sample of 54,777 stars. We use Fe to track total metallicity and Mg alone as an $\alpha$-element.

@@

To understand the key events driving the behavior around the dashed line in Figure~\ref{fig:alpha}, we examine the evolution of several properties of our galaxy in Figure~\ref{fig:history}: its SFH, the BH accretion rate, and the growth of the virial mass. The vertical dashed line in each panel marks the transition at $t\sim3.2\,\Gyr$ from the high- to low-$\alpha$ sequences, as in Figure~\ref{fig:alpha}. The upper panel in the left column shows the the SFR (computed for all gas cells in the \edit1{\texttt{SUBFIND} subhalo}). There are two peaks at $t\sim2.5\,\Gyr$ and $t\sim4.5\,\Gyr$ with maximum values of $50\,\Msun/\textrm{yr}$ and $30\,\Msun/\textrm{yr}$, respectively. Around the high- to low-$\alpha$ transition, there is a dip in the SFR, which drops by an order of magnitude to about $3\,\Msun/\textrm{yr}$. The right panel zooms in on the period between $t=2\,\Gyr$ and $5\,\Gyr$, where we observe a sharp recovery in the SFR following the quiescent phase. In the span of a single snapshot (roughly $150\,\Myr$), the SFR increases from about $3\,\Msun/\textrm{yr}$ to $10\,\Msun/\textrm{yr}$.

@@

\edit1{As discussed in Paper~I, the key to generating an $\alpha$-bimodality is halting star formation within specific metallicity ranges. A global quiescent period is a sufficient but not necessary condition for the formation of such gaps. To illustrate this point, we present Figure~\ref{fig:zSFH}, which compares the global SFH in black with the SFH in narrow metallicity bins, color-coded to match the bins shown in Figure~\ref{fig:fig1}. This plot uses the $z=0$ distribution of star particle formation times, resulting in minor differences from the SFH shown in Figure~\ref{fig:history}. Notably, we observe a significant drop in the SFR within these metallicity bins, lasting up to $\sim1\,\Gyr$. However, because the timing of these metallicity-dependent gaps differs, the total SFR never falls below $\sim10\,\Msunyr$.}

\begin{figure}
  \centering
  \includegraphics[width=\columnwidth]{392276_zSFH.pdf}
  \caption{\edit1{Global star formation history (black) compared with the star formation history in narrow metallicity bins, color-coded as in Figure~\ref{fig:fig1}. This plot demonstrates that the key condition for generating an $\alpha$-bimodality, the cessation of star formation in specific metallicity ranges, is satisfied. The metallicity-dependent SFR drops nearly to zero in every bin for periods extending up to $\sim1\,\Gyr$. The timing and duration of metallicity-dependent gaps can vary, preventing the total SFR from falling below $\sim10\,\Msunyr$.}}
  \label{fig:zSFH}
\end{figure}

@@

This figure shows the following sequence of events:
\begin{enumerate}
    \item Bar forms: A steady increase in the bar strength, as indicated by $A_2/A_0$ for star particles with $R<2\,\kpc$, from $\sim0.05$ to $0.4$ starting around $2.8\,\textrm{Gyr}$. This rise is accompanied by the appearance of elongated features in the gas and stars consistent with a bar.
    \item BH accretion increases: Following the increase in bar strength by about a snapshot ($\sim150\,\Myr$ here), the BH accretion rate ($\dot{M}_{\textrm{BH}}/\dot{M}_{\textrm{edd}}$) shows a significant spike, rising from a minimum of $\sim0.08$ at $t=2.84\,\Gyr$ to a maximum of $\sim0.29$ at $t=3.13\,\Gyr$ for one snapshot.
    \item SFR declines: The SFR declines starting from a maximum of $54.3\,\Msunyr$ at $t=2.84\,\Gyr$ down to $3.6\,\Msunyr$ at $t=3.28\,\Gyr$. In the next snapshot at $t=3.45\,\Gyr$ the SFR recovers to $12.2\,\Msunyr$. Figure~\ref{fig:history} shows that it reaches its second maximum of $30.9\,\Msunyr$ at $4.5\,\Gyr$. \edit1{We show in Figure~\ref{fig:zSFH} that the global SF suppression is coincident with a much more dramatic metallicity dependent SF gap, with the SFR dropping nearly to $0\,\Msunyrdex$ in several bins.}
\end{enumerate}

@@

\subsection{Cause of Bimodality}\label{ssec:bim_cause}
We argue that the available evidence for the cause of the bimodality aligns with the scenario outlined in Paper~I. That study presented an idealized simulation resembling the merger between the Milky Way and GSE, where the orbital parameters were varied slightly across a grid of 27 simulations. The results showed that simulations featuring a brief quiescent period produced a bimodal abundance distribution. Two conditions for the bimodality to arise have to be met: (1) a declining \alphaFe{} with time (here done through our post-processing step, see Figure~\ref{fig:alpha}), and (2) a gap in \edit1{the metallicity-dependent} SFR. In Paper~I, the gap in SFR was due to a merger, but here we argue it is due to the formation of a bar.

@@

\edit1{This global quiescent period is symptomatic of a more dramatic reduction in the star formation rate in narrow metallicity bins, which almost entirely vanishes (Figure~\ref{fig:zSFH}). As we showed in Paper~I, it is the metallicity-dependent quiescence in the presence of a declining gas-phase \MgFe{} -- not global quiescence -- that leads to an $\alpha$-bimodality. In the fiducial TNG model, the \MgFe{} reduction with time is not rapid enough, leading to the post-processing discussed in Section~\ref{ssec:sfe}. In the real universe, a global quiescent period of a duration of $\sim300\,\Myr$ would lead to a reduction in enrichment from Type~II relative to Type~Ia SNe leads to a lower rate of Mg production. The typical lifetime of a Type~II SN progenitor in this model is $\sim40\,\Myr$ \citep{2018MNRAS.473.4077P}, and so the hundreds of Myr of suppressed SF would be short enough to restrict the production of $\alpha$-elements.}

@@

Notably, both the fiducial and $\alpha$-enhanced galaxy show a slight rebound effect in \MgFe{}. The star particles which form directly after the dashed line when the SFR has just recovered have a slightly lower \MgFe{} than stars which form later on, by about $0.1$ to $0.2\,\dex$. \edit1{We argue that it is plausible that} during the period of suppressed star formation, the \alphaFe{} ratio of the star-forming gas drops sharply due to the reduced contribution of Type~II relative to Type~Ia SNe. Later, the \alphaFe{} of the gas will recover when the SFR also recovers, but there is a brief window when old, low-$\alpha$ stars can form. A similar behavior was seen in the one-zone models with bursty SFHs in \citet{2020MNRAS.498.1364J}.

@@

\subsection{Cause of Quiescence}\label{ssec:cause_qui}
In Paper~I, AGN activity from a merger was the suspected cause for the quiescent period. In our galaxy here, there is indeed a brief burst in AGN accretion at the time of the merger (middle panel of Figure~\ref{fig:history}). Based on this burst, it is also reasonable to suspect that AGN activity is also responsible for the quiescent period in our galaxy, \edit1{noting that the real cause of the formation of the $\alpha$-bimodality is a metallicity-dependent quiescent period (see Paper~I and Figure~\ref{fig:zSFH})}. However, we argued in Section~\ref{ssec:sequence_of_events} that the cause of the localized spike in the BH accretion rate is not due to a merger but instead due to the formation/strengthening of a bar. \edit1{This connection is further supported by estimates that the Milky Way's bar formed $\sim8\,\Gyr$ ago \citep[e.g.,][]{2019MNRAS.490.4740B,2024MNRAS.530.2972S}, roughly concurrent with the formation of the bimodality \citep{2013A&A...560A.109H,2023MNRAS.525.2208R,2024MNRAS.535..392L}.}

@@

The formation of the bimodality, when the galaxy transitions from producing high- to low-\MgFe{} star particles (Figure~\ref{fig:alpha}), is coincident with \edit1{both a global and metallicity-dependent suppression of star formation (Figures~\ref{fig:history} and \ref{fig:zSFH})}. This suppression of star formation is preceded by the formation of a bar and subsequent AGN activity (Figure~\ref{fig:seq}). \edit1{This scenario is plausible for the Milky Way, as recent estimates indicate that the bar formed around $8\,\Gyr$ ago \citep{2019MNRAS.490.4740B,2024MNRAS.530.2972S}, coinciding with the onset of the bimodality.}

\edit1{The lack of star formation in a narrow metallicity bin in the presence of a decline in \MgFe{} naturally leads to a gap in the present-day distribution of \MgFe{} (see Paper~I). However, in the fiducial TNG  model, the decline in \MgFe{} is not rapid enough during the period of quiescence to produce an $\alpha$-bimodality. Our post-processing $\alpha$-enhancement artificially injects this declination. In the real universe, a global reduction in the SFR may expedite the drop in \MgFe{} since the number of Type~II relative to Type~Ia SNe would drop.}

This work adds further support to a scenario in which a quiescent period in the Milky Way's past is a plausible explanation for the Milky Way's abundance bimodality. We argued in Paper~I that the GSE merger could trigger this period. In this work we have argued that the formation of the Milky Way's bar could be responsible. Regardless of this scenario's relevance to the Milky Way, we also predict that the presence of a bar in external galaxies is correlated with the presence of an $\alpha$-bimodality. \edit1{Because of observational errors in age, it is not possible to make a direct comparison between the simulated galaxy and dataset (Figure~\ref{fig:alpha} and Section~\ref{ssec:compare_obs}). In the future, more numerous and precise age estimates of old, metal-poor stars may be able to distinguish the formation scenarios of the $\alpha$-bimodality.}

@@

\figsetstart
\figsetnum{8}
\figsettitle{Abundance plane for a random subset of galaxies in our Milky Way-progenitor mass sample, with and without $\alpha$-enhancement.}
\figsetgrpstart
\figsetgrpnum{8.1}
\figsetgrptitle{Subhalo 392276.}
\figsetplot{app_392276.pdf}
\figsetgrpnote{Abundance plane for subhalo 392276.}
\figsetgrpend
\figsetgrpstart
\figsetgrpnum{8.2}
\figsetgrptitle{Subhalo 2.}
\figsetplot{app_2.pdf}
\figsetgrpnote{Abundance plane for subhalo 2.}
\figsetgrpend
\figsetgrpstart
\figsetgrpnum{8.3}
\figsetgrptitle{Subhalo 10.}
\figsetplot{app_10.pdf}
\figsetgrpnote{Abundance plane for subhalo 10.}
\figsetgrpend
\figsetgrpstart
\figsetgrpnum{8.4}
\figsetgrptitle{Subhalo 143882.}
\figsetplot{app_143882.pdf}
\figsetgrpnote{Abundance plane for subhalo 143882.}
\figsetgrpend
\figsetgrpstart
\figsetgrpnum{8.5}
\figsetgrptitle{Subhalo 167392.}
\figsetplot{app_167392.pdf}
\figsetgrpnote{Abundance plane for subhalo 167392.}
\figsetgrpend
\figsetgrpstart
\figsetgrpnum{8.6}
\figsetgrptitle{Subhalo 289388.}
\figsetplot{app_289388.pdf}
\figsetgrpnote{Abundance plane for subhalo 289388.}
\figsetgrpend
\figsetgrpstart
\figsetgrpnum{8.7}
\figsetgrptitle{Subhalo 300903.}
\figsetplot{app_300903.pdf}
\figsetgrpnote{Abundance plane for subhalo 300903.}
\figsetgrpend
\figsetgrpstart
\figsetgrpnum{8.8}
\figsetgrptitle{Subhalo 348901.}
\figsetplot{app_348901.pdf}
\figsetgrpnote{Abundance plane for subhalo 348901.}
\figsetgrpend
\figsetgrpstart
\figsetgrpnum{8.9}
\figsetgrptitle{Subhalo 398784.}
\figsetplot{app_398784.pdf}
\figsetgrpnote{Abundance plane for subhalo 398784.}
\figsetgrpend
\figsetgrpstart
\figsetgrpnum{8.10}
\figsetgrptitle{Subhalo 404818.}
\figsetplot{app_404818.pdf}
\figsetgrpnote{Abundance plane for subhalo 404818.}
\figsetgrpend
\figsetgrpstart
\figsetgrpnum{8.11}
\figsetgrptitle{Subhalo 425719.}
\figsetplot{app_425719.pdf}
\figsetgrpnote{Abundance plane for subhalo 425719.}
\figsetgrpend
\figsetgrpstart
\figsetgrpnum{8.12}
\figsetgrptitle{Subhalo 439099.}
\figsetplot{app_439099.pdf}
\figsetgrpnote{Abundance plane for subhalo 439099.}
\figsetgrpend
\figsetgrpstart
\figsetgrpnum{8.13}
\figsetgrptitle{Subhalo 465255.}
\figsetplot{app_465255.pdf}
\figsetgrpnote{Abundance plane for subhalo 465255.}
\figsetgrpend
\figsetgrpstart
\figsetgrpnum{8.14}
\figsetgrptitle{Subhalo 494709.}
\figsetplot{app_494709.pdf}
\figsetgrpnote{Abundance plane for subhalo 494709.}
\figsetgrpend
\figsetgrpstart
\figsetgrpnum{8.15}
\figsetgrptitle{Subhalo 510273.}
\figsetplot{app_510273.pdf}
\figsetgrpnote{Abundance plane for subhalo 510273.}
\figsetgrpend
\figsetgrpstart
\figsetgrpnum{8.16}
\figsetgrptitle{Subhalo 547293.}
\figsetplot{app_547293.pdf}
\figsetgrpnote{Abundance plane for subhalo 547293.}
\figsetgrpend
\figsetgrpstart
\figsetgrpnum{8.17}
\figsetgrptitle{Subhalo 576516.}
\figsetplot{app_576516.pdf}
\figsetgrpnote{Abundance plane for subhalo 576516.}
\figsetgrpend
\figsetend
